---
title: "Bridging the Credit Gap: A Framework for Implementing Alternative Credit Scoring in South Africa"
subtitle: "ECO5016W - Minor Dissertation in FinTech (Non-unitary option)"
author: "Cesaire Tobias"
date: "2025-11-01"
format:
  pdf:
    date-format: "DD MMMM, YYYY"
    toc: true
    toc-depth: 3
    lot: false
    lof: true
    number-sections: true
    colorlinks: true
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.5
    cite-method: citeproc
    bibliography: references.bib
    csl: harvard-cite-them-right.csl
    link-citations: true
    keep-tex: false
    fig-pos: 'H'
    tbl-pos: 'H'
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{placeins}
        \usepackage{array}
        \usepackage{booktabs}
        \floatplacement{figure}{H}
        \floatplacement{table}{H}
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com/"))

required_packages <- c("dplyr", "ggplot2", "stringr", "tidyr", "forcats", "flextable", "officer", "knitr", "scales", "rsvg")
optional_packages <- c("cqThemes")

github_packages <- list(
  cqThemes = "ces0491/cqThemes"
)

if (!"remotes" %in% installed.packages()[, "Package"]) {
  install.packages("remotes", quiet = TRUE)
}

installed <- installed.packages()[, "Package"]
missing_packages <- setdiff(required_packages, installed)

cran_missing <- setdiff(missing_packages, names(github_packages))
if (length(cran_missing) > 0) {
  install.packages(cran_missing, quiet = TRUE)
}

for (pkg in optional_packages) {
  if (!pkg %in% installed && pkg %in% names(github_packages)) {
    tryCatch({
      remotes::install_github(github_packages[[pkg]], quiet = TRUE)
    }, error = function(e) NULL)
  }
}

invisible(lapply(required_packages, function(pkg) {
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}))

for (pkg in optional_packages) {
  tryCatch({
    suppressPackageStartupMessages(library(pkg, character.only = TRUE))
  }, error = function(e) NULL)
}

cqthemes_available <- "cqThemes" %in% (.packages())

if (cqthemes_available) {
  tryCatch({
    use_pdf_theme(seed = 0491)
  }, error = function(e) {
    cqthemes_available <<- FALSE
  })
}

if (!cqthemes_available) {
  theme_set(theme_minimal())
}

if (cqthemes_available) {
  get_colors <- function(n = 4, type = "categorical") {
    cq_colors(n, type = type)
  }
  get_theme <- function() theme_cq_pdf()
} else {
  get_colors <- function(n = 4, type = "categorical") {
    colors <- c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", "#8c564b")
    return(colors[1:min(n, length(colors))])
  }
  get_theme <- function() theme_minimal()
}

options(scipen = 999)
```

```{r wordcount, include=FALSE}
count_words_qmd <- function(file_path) {
  raw_text <- readLines(file_path, warn = FALSE)
  
  body_text <- ""
  in_code_chunk <- FALSE
  in_yaml <- FALSE
  yaml_delim_count <- 0
  
  for (i in seq_along(raw_text)) {
    line <- raw_text[i]
    line_trimmed <- trimws(line)
    
    if (line_trimmed == "---") {
      yaml_delim_count <- yaml_delim_count + 1
      if (yaml_delim_count == 1) {
        in_yaml <- TRUE
        next
      } else if (yaml_delim_count == 2) {
        in_yaml <- FALSE
        next
      }
    }
    
    if (in_yaml) next
    
    if (grepl("^#+\\s*(References|Bibliography|Works Cited|Literature Cited)", 
              line_trimmed, ignore.case = TRUE)) {
      break
    }
    
    if (grepl("^```", line_trimmed)) {
      in_code_chunk <- !in_code_chunk
      next
    }
    
    if (in_code_chunk) next
    
    if (grepl("^#+", line_trimmed)) next
    
    if (grepl("^fig\\.|^tbl\\.|fig\\.cap|caption\\s*=|include\\s*=|echo\\s*=", line_trimmed)) next
    
    processed_line <- line
    
    processed_line <- gsub("`[^`]*`", "", processed_line)
    processed_line <- gsub("\\*\\*([^*]+)\\*\\*", "\\1", processed_line)
    processed_line <- gsub("\\*([^*]+)\\*", "\\1", processed_line)
    processed_line <- gsub("_([^_]+)_", "\\1", processed_line)
    processed_line <- gsub("\\[([^]]+)\\]\\([^)]+\\)", "\\1", processed_line)
    processed_line <- gsub("!\\[([^]]*)\\]\\([^)]+\\)", "", processed_line)
    processed_line <- gsub("[@][A-Za-z0-9_:-]+", "", processed_line)
    processed_line <- gsub("<[^>]+>", "", processed_line)
    
    body_text <- paste(body_text, processed_line)
  }
  
  body_text <- gsub("[[:punct:]]+", " ", body_text)
  words <- unlist(strsplit(body_text, "\\s+"))
  words <- words[nchar(trimws(words)) > 0]
  
  return(length(words))
}

word_count <- count_words_qmd(knitr::current_input())
```

```{r svg-conversion, include=FALSE}
# SVG to PNG Conversion Function
convert_svg_to_png <- function(svg_path, png_path = NULL, dpi = 300) {
  # If no PNG path specified, replace .svg with .png
  if (is.null(png_path)) {
    png_path <- sub("\\.svg$", ".png", svg_path)
  }
  
  # Check if SVG file exists
  if (!file.exists(svg_path)) {
    warning(paste("SVG file not found:", svg_path))
    return(FALSE)
  }
  
  # Only convert if SVG is newer than PNG (or PNG doesn't exist)
  if (!file.exists(png_path) || file.mtime(svg_path) > file.mtime(png_path)) {
    tryCatch({
      # Method 1: Try using rsvg package (most reliable)
      if (requireNamespace("rsvg", quietly = TRUE)) {
        bitmap <- rsvg::rsvg(svg_path, width = dpi * 10)
        png::writePNG(bitmap, png_path)
        message(paste("Converted", basename(svg_path), "to PNG using rsvg"))
        return(TRUE)
      }
      
      # Method 2: Try system command with Inkscape (if available)
      inkscape_test <- suppressWarnings(system("inkscape --version", intern = FALSE))
      if (inkscape_test == 0) {
        cmd <- sprintf("inkscape %s --export-filename=%s --export-dpi=%d", 
                      svg_path, png_path, dpi)
        system(cmd, ignore.stdout = TRUE, ignore.stderr = TRUE)
        if (file.exists(png_path)) {
          message(paste("Converted", basename(svg_path), "to PNG using Inkscape"))
          return(TRUE)
        }
      }
      
      # Method 3: Try rsvg-convert (Linux/Mac)
      rsvg_test <- suppressWarnings(system("which rsvg-convert", intern = FALSE))
      if (rsvg_test == 0) {
        cmd <- sprintf("rsvg-convert -d %d -p %d %s -o %s", 
                      dpi, dpi, svg_path, png_path)
        system(cmd, ignore.stdout = TRUE, ignore.stderr = TRUE)
        if (file.exists(png_path)) {
          message(paste("Converted", basename(svg_path), "to PNG using rsvg-convert"))
          return(TRUE)
        }
      }
      
      # Method 4: Try ImageMagick convert
      convert_test <- suppressWarnings(system("which convert", intern = FALSE))
      if (convert_test == 0) {
        cmd <- sprintf("convert -density %d %s %s", dpi, svg_path, png_path)
        system(cmd, ignore.stdout = TRUE, ignore.stderr = TRUE)
        if (file.exists(png_path)) {
          message(paste("Converted", basename(svg_path), "to PNG using ImageMagick"))
          return(TRUE)
        }
      }
      
      warning(paste("Could not convert SVG. Please install rsvg package or Inkscape/ImageMagick"))
      return(FALSE)
      
    }, error = function(e) {
      warning(paste("Error converting SVG:", e$message))
      return(FALSE)
    })
  } else {
    message(paste("PNG is up-to-date for", basename(svg_path)))
    return(TRUE)
  }
}

# Convert all SVG files in the directory
svg_files <- list.files(pattern = "\\.svg$", full.names = TRUE)

for (svg_file in svg_files) {
  convert_svg_to_png(svg_file)
}

# Ensure diagram is converted
main_diagram_svg <- "acs-framework-diagram.svg"
main_diagram_png <- "acs-framework-diagram.png"

if (file.exists(main_diagram_svg)) {
  conversion_success <- convert_svg_to_png(main_diagram_svg, main_diagram_png)
  
  # Create a flag for the document to know which format to use
  diagram_format <- ifelse(conversion_success && file.exists(main_diagram_png), 
                          "png", "svg")
  diagram_file <- ifelse(diagram_format == "png", 
                         main_diagram_png, main_diagram_svg)
  
  # Store in options for later use
  options(diagram_file = diagram_file)
  options(diagram_format = diagram_format)
  
  message(paste("Will use", diagram_format, "format for main diagram"))
} else {
  warning("Main diagram SVG file not found!")
  options(diagram_file = "acs-framework-diagram.svg")
  options(diagram_format = "svg")
}
```

\newpage

# Abstract {.unnumbered}

South Africa faces an economically stifling financial inclusion challenge: an estimated 16 million adults (41% of working age adults [@statista2024population]) remain excluded from formal credit markets due to insufficient credit histories [@transunion2025] despite credit bureaus maintaining records for 28.9 million consumers [@ncr2025]. This exclusion constrains economic participation and perpetuates intergenerational poverty [@piraino2015; @arellano2024], narrowing excluded individuals' options towards informal lenders who may charge interest rates of up to 50% per month [@james2015].

To address the challenge, this dissertation presents a conceptual implementation framework for alternative credit scoring in South Africa through systematic literature synthesis and comparative case analysis. The research examines international implementations (Brazil's Cadastro Positivo, Kenya's M-Shwari, India's Account Aggregator) to extract design principles applicable to South African conditions, synthesizing insights from @chioda2025's work on lending to no-credit-history borrowers and @fuster2022's analysis of machine learning in credit markets.

The proposed framework centers on three integrated components. First, a hybrid federated-centralized data architecture that could resolve tensions between processing efficiency and privacy protection under POPIA, utilizing concepts from edge computing and privacy-preserving technologies. Second, a governance structure establishing an Alternative Credit Scoring Authority within existing regulatory frameworks, implementing mandatory explainability standards and three-tier appeals processes based on @karimi2021's algorithmic recourse principles. Third, an incentive alignment mechanism designed to ensure commercial viability through proposed regulatory capital relief for participating lenders and revenue-sharing for data providers, following @olckers2023's game-theoretic insights.

The implementation roadmap proposes progression through regulatory sandboxes, scaled deployment, and ecosystem integration—leveraging South Africa's rapidly growing Buy Now Pay Later sector as a potential proving ground. Policy recommendations include suggested amendments to the National Credit Act recognizing alternative data sources, proposed POPIA sectoral codes for credit assessment, and industry-wide data sharing standards compatible with emerging open banking initiatives. The framework addresses risks of algorithmic discrimination documented by @barocas2016 while proposing mechanisms to expand meaningful credit access to previously excluded populations.

::: {=latex}
\vspace{1\baselineskip}
:::

**Keywords:** alternative credit scoring, financial inclusion, algorithmic fairness, data governance, POPIA compliance, Buy Now Pay Later, credit invisibles, South Africa, fintech regulation, machine learning in finance

::: {=latex}
\vspace{1\baselineskip}
:::

**Word Count: `r formatC(word_count, format="d", big.mark=",")`**

---

\newpage

# Introduction

South Africa's credit market presents a paradox that undermines both economic efficiency and principles of social justice. While the country maintains Africa's most sophisticated and developed financial sector, with major banks' operations comprising 83% of total banking sector assets [@pwc2025], an estimated 16 million adults remain systematically excluded from formal credit markets due to insufficient or non-existent credit histories [@transunion2025]. According to the National Credit Regulator (NCR), credit bureaus maintained records for 28.9 million credit-active consumers as of March 2025 [@ncr2025], yet analysis by @finmark2024 reveals that 23% of adults remain credit invisible while an additional 15% have thin files insufficient for traditional scoring^[Traditional scoring uses methods such as regression, decision trees, and other statistical analyses to generate a credit score using limited amounts of structured data, such as amount of loan, type of loan, maturity of loan, guarantees and collateral value, historical payment performance such as default information and payments in arrears, amounts owed, length of credit history, new credit, and types of credit [@worldbank2019credit]]. These "credit invisibles"^[Individuals with no credit history or so little that it is insufficient to compile traditional scoring metrics [@transunion2024invisible]] span diverse demographic and economic segments: informal traders operating entirely outside formal financial systems [@fourie2018], gig economy workers navigating precarious employment without traditional income documentation [@hunt2019], recent university graduates entering formal employment but lacking requisite credit histories [@james2015], and vast rural populations geographically and economically distant from traditional financial infrastructure [@wentzel2016].

The macroeconomic consequences of this widespread exclusion extend far beyond individual hardship to constrain entire economic potential. The informal sector represents an estimated 18% of total non-agricultural employment and 6.0% of GDP [@statssa2019], operates largely outside the formal credit system, limiting growth potential and economic formalization [@rogan2019; @skinner2018]. Research by @kingdon2007 demonstrates how credit constraints prevent informal enterprises from scaling operations and transitioning to formal status. At the microeconomic level, excluded individuals face stark choices between foregoing credit-dependent opportunities entirely or resorting to informal lending mechanisms that often charge exploitative rates [@james2015; @mashigo2012], creating debt spirals that entrench poverty across generations [@ardington2015]. 

The broader social costs manifest in severely limited economic mobility, dramatically reduced entrepreneurial activity particularly in township economies [@charman2017], and the perpetuation of inequality in a society already grappling with extreme income disparities [@zizzamia2020; @schotte2018]. As @piraino2015 estimates, South Africa exhibits an intergenerational earnings elasticity of 0.6, meaning children's economic outcomes remain strongly tied to their parents' position. Interestingly, research from the United States by @arellano2024 demonstrates complex relationships between credit access and intergenerational mobility: while a 10% increase in parents' unused revolving credit is associated with 0.28% to 0.37% higher earnings for their children, the overall democratization of credit between the 1970s and 2000s actually reduced intergenerational mobility due to offsetting effects of bankruptcy policy on low-income household savings and investment behaviors. This paradox underscores the importance of not just expanding credit access, but doing so through carefully designed mechanisms that genuinely enhance economic opportunity.

Alternative credit scoring emerges as a potential solution to this structural exclusion. By leveraging non-traditional data sources—including mobile phone usage patterns [@bjorkegren2019], utility payment histories [@turner2012], digital transaction records [@jagtiani2019], social media activity [@wei2016], and psychometric assessments [@arraiz2017]—combined with sophisticated machine learning techniques [@khandani2010], lenders could theoretically assess creditworthiness for previously unscorable populations with accuracy approaching or exceeding traditional methods. The seminal work by @berg2020 demonstrates that digital footprints contain significant information content for predicting consumer default, achieving predictive accuracy comparable to traditional credit bureau scores. Similar findings by @gambacorta2019 using data from Chinese fintech platforms confirm the predictive power of alternative data in emerging markets. The proliferation of Buy Now Pay Later (BNPL) providers in South Africa, with the market reaching US\$717.3 million in 2024 and projected to hit US$820 million by end of 2025 [@easyequities2025], demonstrates both substantial market appetite and technical feasibility for alternative credit assessment approaches, though @wessels2021 cautions about potential over-indebtedness risks.

However, implementation raises multifaceted challenges inadequately addressed by existing academic literature and policy frameworks. While developed market studies by @fuster2022 examining U.S. mortgage applications and @dimaggio2022 analyzing fintech lending to "invisible primes" focus primarily on marginal improvements to mature credit systems with established infrastructure, South Africa requires fundamental infrastructure development from the ground up [@schoeman2021]. The challenge encompasses not merely technical implementation of algorithms and data pipelines [@baesens2016], but complex regulatory alignment under the Protection of Personal Information Act (POPIA), where tensions between data minimization requirements and big data analytics create significant compliance challenges [@vanderbank2020; @botha2021; @davids2022]. Implementation must navigate delicate stakeholder coordination across historically fragmented data ecosystems—with four major credit bureaus, multiple telecommunications providers, and numerous utilities operating incompatible systems [@groepe2017; @cenfri2019; @finmark2020]—while addressing profound ethical considerations in a society where apartheid's spatial legacy still determines credit access patterns [@francis2019; @levy2021] and historical exclusions from property ownership continue to affect creditworthiness assessments [@chatterjee2022; @bond2013].

@chioda2025 provide evidence that fintech lenders can successfully serve no-credit-history borrowers while maintaining portfolio quality comparable to traditional lending. Their analysis of millions of loans across multiple emerging markets demonstrates that alternative data combined with appropriate risk management frameworks can expand access without compromising financial stability. This finding is particularly relevant for South Africa, where concerns about systemic risk have historically constrained innovation in credit markets.

This dissertation addresses a gap in both academic literature and policy discourse: how can South Africa develop and implement an alternative credit scoring system that expands meaningful financial inclusion while simultaneously maintaining appropriate consumer protections and ensuring commercial viability for market participants? The research develops a conceptual framework through literature synthesis and comparative analysis, proposing concrete mechanisms specifically suitable for South African conditions, regulatory requirements, and social imperatives.

The framework systematically integrates three essential dimensions required for successful implementation. First, technical architecture for responsible data aggregation and model deployment, carefully considering infrastructure constraints and capacity limitations. Second, governance mechanisms ensuring algorithmic accountability and fairness within existing regulatory frameworks while allowing for innovation and experimentation. Third, economic incentives properly aligning diverse stakeholder interests toward sustainable, scalable implementation rather than pilot projects that fail to achieve systemic impact.

# Literature Review

## Theoretical Foundations of Credit Scoring

The theoretical foundations of credit scoring trace back to @durand1941's pioneering application of discriminant analysis to consumer loan assessment, establishing the principle that statistical methods could predict repayment probability more accurately than subjective judgment. This quantitative revolution gained substantial momentum through @altman1968's influential Z-score model, which demonstrated that carefully selected financial ratios could predict corporate bankruptcy with high accuracy. These early innovations established the paradigm that objective, data-driven assessment could significantly outperform traditional relationship-based lending decisions.

However, @mester1997 identifies limitations inherent in traditional scoring approaches that become particularly acute in developing market contexts. Traditional models require extensive historical data spanning multiple economic cycles, assume stable relationships between predictors and outcomes that may not hold during structural economic shifts, and exhibit consistently poor performance for thin-file applicants who lack established credit histories — precisely those populations excluded in emerging markets. These limitations create what scholars term the "credit scoring paradox": those most in need of credit to build economic stability are systematically excluded by the very mechanisms designed to assess creditworthiness.

The information asymmetry framework advanced by @stiglitz1981 provides the theoretical grounding for understanding persistent credit market failures in developing economies. In contexts of imperfect information where lenders cannot effectively distinguish between high and low-risk borrowers, rational lender behavior leads to adverse selection and credit rationing even at market-clearing interest rates. @jaffee1976 demonstrate through formal modeling that this results in equilibrium credit exclusion where creditworthy individuals are denied access not due to their actual risk profile but due to informational opacity. South African manifestations of these theoretical predictions are evident in the credit market structure, where significant portions of the population remain excluded from formal credit [@transunion2025] despite potential creditworthiness.

## Alternative Data and Machine Learning Applications

The alternative data literature has evolved into three distinct but interconnected streams examining different data sources and their predictive power for credit assessment. The first stream, exemplified by recent work including @chioda2025, focuses on structured alternative data including telecommunications records, utility payments, and digital transaction histories. @chioda2025's analysis of millions of loans to no-credit-history borrowers across Latin America and Africa demonstrates that fintech lenders using alternative data achieve default rates comparable to traditional lending while serving previously excluded populations. These studies consistently find alternative data significantly improves prediction accuracy for thin-file applicants while maintaining comparable performance for traditional segments.

The second stream explores unstructured and behavioral data sources with increasing sophistication. Mobile phone metadata—call patterns, recharge frequencies, location stability — proves particularly valuable in emerging markets. @jack2016 demonstrate through Kenya's M-Shwari platform how mobile money data enables credit extension to previously unbanked populations while maintaining portfolio quality. E-commerce browsing patterns reveal behavioral signatures of financial responsibility: comparison shopping indicates prudence while impulsive purchasing patterns correlate with elevated default risk. Psychometric assessments, while less studied in African contexts, show promise in capturing personality traits and cognitive patterns that traditional financial metrics miss.

The third stream, advanced by @fuster2022, addresses machine learning applications to credit scoring with particular attention to distributional effects. Their analysis of U.S. mortgage applications reveals a paradox: while machine learning models improve overall prediction accuracy compared to traditional logistic regression, they can simultaneously exacerbate disparities in approval rates across demographic groups. @dimaggio2022 identify "invisible primes" — creditworthy borrowers undetectable by traditional scoring but identifiable through alternative data — demonstrating vast untapped market potential. However, these advances come with challenges: complex models become "black boxes" violating regulatory requirements for explanation, while simpler interpretable models sacrifice predictive power.

## Algorithmic Fairness and Discrimination

The algorithmic fairness literature reveals tensions in applying machine learning to credit decisions with profound implications for financial inclusion. @barocas2016's seminal work establishes that algorithms can perpetuate and amplify discrimination through multiple mechanisms: biased training data reflecting historical inequalities, seemingly neutral variables that inadvertently proxy for protected characteristics, and feedback loops that amplify initial disparities over time. Their analysis demonstrates how facially neutral algorithms can produce discriminatory outcomes even without explicit use of protected characteristics.

@kleinberg2017 formalize the mathematical impossibility of simultaneously satisfying different fairness definitions—demographic parity, equalized odds, and calibration—except under unrealistic conditions of equal base rates across groups. This impossibility theorem has considerable implications for credit scoring in diverse societies like South Africa, where historical inequalities create vastly different risk distributions across demographic groups.

The concept of algorithmic recourse, advanced by @karimi2021, offers a potential path forward by shifting focus from static fairness to dynamic improvement. Their framework demonstrates that providing specific, actionable guidance to rejected applicants—rather than opaque rejections—can significantly increase subsequent approval rates. This approach transforms credit denial from a dead end into a developmental process, particularly crucial for first-time borrowers.

@olckers2023 extend this analysis by examining incentives for offering algorithmic recourse from a game-theoretic perspective. Their model reveals conditions under which lenders benefit from helping rejected applicants become creditworthy, creating mutual value through expanded customer bases and improved risk profiles. This work is particularly relevant for South Africa, where large excluded populations represent untapped market potential rather than inherently high-risk segments.

## Regulatory Frameworks and International Experience

The regulatory literature reveals divergent global approaches to governing alternative credit scoring, each with distinct implications for innovation and inclusion. The European Union's General Data Protection Regulation (GDPR) Article 22 grants individuals rights not to be subject to purely automated decision-making with significant effects, though @kaminski2019 argues meaningful human review proves impractical at the scale required for mass market credit. South Africa's Protection of Personal Information Act (POPIA) creates similar tensions between data minimization principles and machine learning's requirement for comprehensive data analysis.

International experiences offer valuable lessons requiring careful contextual adaptation. Brazil's Cadastro Positivo system achieved remarkable success with millions enrolled through mandatory participation and standardized data sharing protocols [@bcb2024]. Kenya's M-Shwari platform demonstrates how mobile money integration can extend credit to previously unbanked individuals while maintaining portfolio quality [@jack2016]. India's Account Aggregator framework creates consent-based data sharing infrastructure that balances privacy with inclusion imperatives.

However, regulatory transferability remains limited by institutional contexts and market structures. South Africa's unique combination of sophisticated financial infrastructure, high mobile penetration, strong regulatory frameworks, and persistent inequality requires tailored approaches rather than wholesale adoption of international models.

# Methodology

## Research Design and Approach

This dissertation employs a conceptual framework development methodology grounded in systematic literature synthesis and comparative case analysis to propose an implementation model for alternative credit scoring in South Africa. The approach follows @hevner2004's design science research principles, creating prescriptive knowledge through the iterative development of artifacts (in this case, a conceptual framework) that address identified problems. Rather than empirical data collection and analysis, this research synthesizes existing knowledge to construct a novel framework tailored to South African conditions.

The research design integrates three methodological components. First, systematic literature review identifies theoretical foundations, empirical evidence, and implementation patterns from academic and grey literature. Second, comparative case analysis of international implementations (Brazil's Cadastro Positivo, Kenya's M-Shwari, and India's Account Aggregator) extracts design principles and lessons applicable to South Africa. Third, conceptual synthesis integrates these insights with South African contextual factors to develop the proposed framework. This approach aligns with @jaakkola2020's conceptual paper methodology, which emphasizes theory synthesis and framework development over empirical testing.

## Literature Review and Synthesis Method

The literature review follows a systematic approach to identify relevant sources while maintaining feasibility within dissertation constraints. Initial searches in Google Scholar, JSTOR, and SSRN used keyword combinations including "alternative credit scoring," "financial inclusion," "algorithmic fairness," and "South Africa" with Boolean operators. Citation tracking expanded the corpus through backward (examining references) and forward (identifying citing papers) searches.

Source selection criteria prioritized relevance and recency: empirical studies from emerging markets received priority over developed market analyses, papers published after 2018 ensured currency given rapid technological evolution, and regulatory documents and industry reports complemented academic sources for practical grounding. The final corpus comprises 63 academic papers, 12 regulatory documents, and 8 industry reports forming the evidence base for framework development.

Thematic analysis identified recurring patterns across sources. Key themes emerged including data architecture choices (centralized versus federated), governance mechanisms (regulatory versus self-regulatory), algorithmic trade-offs (accuracy versus explainability), and implementation challenges (technical, regulatory, social). These themes structure both the literature review presentation and subsequent framework development.

## Comparative Case Analysis

Three international cases provide comparative insights for framework development, selected using @yin2018's replication logic to maximize learning potential. Brazil's Cadastro Positivo exemplifies mandatory participation achieving rapid scale through regulatory coordination. Kenya's M-Shwari demonstrates mobile-first architecture leveraging existing digital infrastructure. India's Account Aggregator illustrates consent-based data sharing balancing privacy with inclusion imperatives.

Case analysis follows structured protocol examining institutional context (regulatory environment, market structure, technological infrastructure), design choices (technical architecture, governance mechanisms, business models), implementation process (phasing, stakeholder engagement, challenge resolution), and outcomes (inclusion metrics, risk performance, unintended consequences). While comprehensive primary data collection exceeded scope constraints, published analyses from @bcb2024, @jack2016, and @sahamati2023 provide sufficient detail for extracting transferable insights.

## Framework Development Process

The conceptual framework emerges through systematic synthesis rather than empirical derivation. Following @jaakkola2020's typology, this represents a "theory synthesis" approach combining multiple theoretical perspectives into an integrated framework. The development process proceeded through three stages.

First, component identification extracted key system elements from literature and cases: technical architecture requirements, governance structures, incentive mechanisms, and implementation processes. Second, relationship mapping connected components through logical dependencies and feedback loops, creating the framework's structural skeleton. Third, contextual adaptation modified generic components for South African specificities including POPIA requirements, NCR regulations, market structure, and infrastructure constraints.

The framework's validity derives from theoretical grounding rather than empirical testing. Each component links to established theory: technical architecture draws on distributed systems and privacy-preserving computation literature, governance mechanisms build on responsive regulation and algorithmic accountability research, and implementation roadmap follows innovation diffusion - following @rogers2003's innovation diffusion principles - and change management principles. This theoretical triangulation provides confidence despite absence of empirical validation.

## Limitations and Scope

This methodology's limitations require explicit acknowledgment. The conceptual nature precludes empirical validation of framework effectiveness. Literature synthesis may miss unpublished insights from practitioners. International case comparisons rely on secondary sources potentially missing implementation nuances. The framework remains untested in practice, requiring pilot implementation for validation.

However, these limitations align with the dissertation's scope as an initial conceptual contribution rather than empirical study. The framework provides structured thinking about alternative credit scoring implementation, identifying key considerations and trade-offs. Future research should empirically test framework propositions through pilot implementations, stakeholder surveys, and quantitative impact assessment. This dissertation establishes the conceptual foundation upon which such empirical work can build.

# Proposed Framework: System Design for Alternative Credit Scoring

```{r figure-setup, echo=FALSE, results='asis'}
# Dynamically determine which file to use based on conversion success
diagram_file <- getOption("diagram_file", "acs-framework-diagram.svg")
diagram_format <- getOption("diagram_format", "svg")

# Check if file actually exists
if (!file.exists(diagram_file)) {
  # Try PNG first, then SVG
  if (file.exists("acs-framework-diagram.png")) {
    diagram_file <- "acs-framework-diagram.png"
  } else if (file.exists("acs-framework-diagram.svg")) {
    diagram_file <- "acs-framework-diagram.svg"
  } else {
    warning("No diagram file found!")
  }
}

# Generate the markdown for the figure
cat(sprintf('![Alternative Credit Scoring Framework: Integrated System Architecture](%s){#fig-framework width=100%%}\n', 
            diagram_file))
```

The proposed framework, illustrated in Figure @fig-framework above, integrates technical, governance, and implementation components into a cohesive system designed for South African conditions. The visual representation demonstrates how the four-layer data architecture connects to the credit application flow, operates under the oversight of the proposed Alternative Credit Scoring Authority (ACSA), and implements consumer protection through a three-tier appeals process. This integrated approach balances innovation with regulation, efficiency with privacy, and inclusion with risk management.

## Technical Architecture for Alternative Credit Scoring

### Data Layer Architecture

The proposed hybrid federated-centralized architecture addresses South Africa's fragmented data landscape while respecting POPIA requirements and commercial sensitivities. This design, drawing from distributed systems theory [@yang2019; @satyanarayanan2017] and privacy-preserving computation literature [@dwork2014], proposes a solution that could balance multiple objectives based on international best practices.

The architecture conceptually comprises four integrated layers, as shown in the left panel of Figure @fig-framework. The data source layer would maintain original data custody, with telecommunications providers retaining call detail records, utilities preserving payment histories, and financial institutions controlling transaction data. Each data custodian could implement standardized APIs following emerging open banking specifications, potentially enabling secure queries without bulk data transfers—theoretically addressing POPIA's data minimization mandate while preserving commercial incentives for participation.

An edge processing layer could deploy lightweight models directly at data sources, performing initial feature extraction and aggregation. For mobile network operators, edge computing [@satyanarayanan2017] might enable real-time feature computation including call pattern stability metrics, recharge frequency indicators, and location diversity scores. This distributed intelligence approach, based on similar implementations documented by @jack2016 in Kenya, could reduce bandwidth requirements while minimizing sensitive data movement across network boundaries.

The orchestration layer would coordinate queries, manage consent, and synthesize insights without storing personal data permanently. When credit applications initiate, the orchestrator could verify consent through secure protocols, query relevant nodes using predefined parameters, and aggregate responses using secure multi-party computation techniques [@yao1982] (a cryptographic method that allows multiple parties to jointly compute a function over their inputs while keeping those inputs private). Only derived features would traverse network boundaries: transaction velocity ratios, payment consistency indices, and behavioral stability scores that reveal creditworthiness without exposing underlying personal information.

International benchmarks suggest such architectures could achieve sub-500ms query latency with appropriate infrastructure investment, though actual performance would depend on local implementation conditions and network capabilities.

### Security and Privacy Framework

The proposed security architecture draws from international data protection best practices, suggesting implementation of defense-in-depth principles. End-to-end encryption using AES-256 could protect data in transit, while homomorphic encryption techniques [@gentry2009] (allowing for analysis without decrypting sensitive information) might enable computation on encrypted data—potentially critical for cross-institutional collaboration given competitive sensitivities. Differential privacy techniques, as documented by @dwork2014, could add calibrated noise preventing individual re-identification while maintaining statistical utility essential for accurate credit assessment.

Zero-knowledge proofs [@goldwasser1989], based on cryptographic protocols described in academic literature, could theoretically enable verification without information disclosure—potentially valuable for income validation in informal sectors where traditional documentation doesn't exist. These protocols might allow proving income exceeds specified thresholds without revealing actual amounts, potentially addressing affordability assessment requirements while protecting privacy.

Implementation timelines and performance characteristics would require validation through pilot testing, as actual deployment depends on technical infrastructure maturity and regulatory approval processes.

## Governance Structure and Oversight Mechanisms

### Institutional Framework Design

The proposed Alternative Credit Scoring Authority (ACSA), illustrated in the governance structure panel of Figure @fig-framework, could operate within the National Credit Regulator structure, leveraging existing regulatory expertise while developing specialized capabilities for algorithmic oversight. This design follows responsive regulation principles described by @ayres1992, potentially creating nested oversight maintaining local autonomy within national standards.

ACSA's proposed mandate could encompass four functions derived from international best practices. First, model certification might evaluate technical performance, fairness metrics, and explainability standards. Second, ongoing monitoring through automated reporting could ensure continued compliance as models evolve. Third, consumer protection mechanisms including mandatory recourse processes and structured appeals procedures. Fourth, innovation facilitation through regulatory sandboxes - following international best practices [@cgap2023] - enabling controlled experimentation.

### Consumer Protection and Recourse

Building on @karimi2021's algorithmic recourse framework and @olckers2023's incentive analysis, the proposed system could implement comprehensive consumer protection beyond traditional credit scoring. Suggested explainability requirements might ensure all decisions include understandable rationales addressing transparency obligations and constitutional administrative justice requirements. Explanations could provide global model behavior, local decision factors, and crucially, counterfactual guidance showing specific steps to achieve approval.

The proposed three-tier appeals process, detailed in the bottom panel of Figure @fig-framework, might balance efficiency with thoroughness. Tier 1 automated reviews could identify clear errors within 24 hours. Tier 2 human reviews might assess borderline cases within seven days. Tier 3 ombudsman appeals could address systemic issues and precedent-setting cases. Published decisions might create jurisprudence guiding future practices.

Behavioral nudges based on choice architecture principles documented by @thaler2008 could guide creditworthiness improvement. Instead of binary rejection, systems might provide specific actionable guidance. Evidence from Kenya's digital lending platforms suggests such feedback mechanisms could improve subsequent approval rates, potentially transforming rejection from dead-end to developmental process.

### Fairness Monitoring and Bias Mitigation

Comprehensive bias detection could employ multiple metrics capturing different fairness dimensions as described in algorithmic fairness literature. Demographic parity might measure approval rate differences across protected groups. Equalized odds could ensure similar true positive and false positive rates across demographics. Calibration might confirm score meanings remain consistent across groups. Combining metrics could provide more comprehensive bias detection than single measures.

Pre-processing techniques documented in machine learning literature could address historical bias in training data through reweighting samples, synthetic minority oversampling, and data augmentation creating counterfactual examples. In-processing constraints might embed fairness directly into model training through adversarial debiasing and multi-objective optimization. Post-processing adjustments could calibrate outputs achieving group fairness while limiting modifications to maintain model integrity.

Proposed fairness thresholds (such as 5% demographic parity difference) represent targets based on international best practices rather than regulatory requirements, requiring stakeholder consensus for implementation.

## Algorithmic Framework and Model Design

### Model Architecture Selection

The proposed ensemble approach could strategically combine multiple algorithms leveraging their respective strengths while mitigating individual weaknesses based on machine learning best practices. Primary prediction might employ gradient boosting machines (a machine learning technique that builds many decision trees sequentially, each correcting the errors of the previous one) for their demonstrated performance on tabular data. Tree-based architecture naturally handles mixed data types, missing values prevalent in thin-files, and non-linear relationships between behavioral patterns and creditworthiness.

Baseline logistic regression models could provide interpretability anchors and performance floors—potentially sufficient for initial screening while maintaining full transparency. When ensemble predictions diverge significantly from baselines, additional scrutiny might ensure decisions remain justifiable to regulators and consumers. Neural network components could capture complex sequential patterns but might comprise limited ensemble weight, activated only for marginal cases where their complexity adds value.

Model performance targets based on international benchmarks suggest potential improvements over traditional scoring, though actual performance would require validation on South African data.

### Feature Engineering Framework

Temporal feature engineering could transform raw time series into predictive signals revealing financial behavior patterns. Rolling statistics over multiple windows might capture behavioral dynamics. Fourier decomposition could identify periodic patterns—weekly salary deposits, monthly obligations—potentially indicating financial stability. Change point detection might identify lifestyle transitions affecting creditworthiness.

Behavioral features derived from transaction data could reveal financial management capabilities invisible to traditional scoring. Domain-specific features might address South African context including township economic indicators, social grant recipient status providing income floor information for millions of beneficiaries, and mobile money usage patterns particularly relevant in underbanked communities.

# Discussion and Policy Implications

## Implementation Roadmap

### Phase 1: Regulatory Sandbox (Months 1-12)

The initial phase could establish a controlled testing environment following international best practices adapted for South African conditions. A sandbox operating under NCR oversight with relaxed regulatory requirements for a limited number of consumers might enable rapid iteration while limiting systemic risk.

Participant selection could ensure diverse perspectives: established banks might provide infrastructure and risk management expertise, microfinance institutions could offer thin-file experience, and fintech companies might contribute technical innovation. This mix could balance stability with innovation potential.

Use case prioritization might focus on maximum learning value with contained risk. Target segments could include first-time borrowers seeking small loans for education or business purposes, informal traders requiring working capital, and young professionals without credit history but stable income. These segments might offer sufficient scale for meaningful validation while limiting systemic impact.

Proposed success metrics could balance multiple objectives: increased approval rates demonstrating inclusion impact, portfolio performance within acceptable ranges of traditional scoring, high explanation satisfaction validating transparency, and fairness metrics within acceptable thresholds across demographics.

### Phase 2: Scaled Deployment (Months 13-30)

Successful sandbox graduates might transition to broader deployment reaching larger consumer segments. Regulatory approval could follow expedited process given sandbox validation. Technical infrastructure might scale through cloud deployment supporting increased volume. Operational processes could mature from manual to automated with human oversight for exceptions.

Interoperability would become critical at scale. Common data standards might enable seamless integration across providers. API specifications published openly could encourage third-party innovation. Consent management might provide user control while enabling data portability. These standards, if mandatory for sandbox graduates, could prevent fragmentation that plagued early mobile money deployments.

Market education might accelerate adoption addressing awareness gaps. Mass media campaigns could explain alternative scoring benefits. Community workshops might provide hands-on assistance. Digital literacy programs integrated with credit education could build capability.

### Phase 3: Ecosystem Integration (Months 31+)

Full integration could transform alternative scoring from parallel system to mainstream infrastructure. Regulatory frameworks might evolve from experimental provisions to permanent regulations. NCR amendments could explicitly recognize alternative data sources and algorithmic decision-making. SARB prudential standards might incorporate alternative scoring into capital calculations.

Proposed success indicators could demonstrate systemic impact: expanded access to previously excluded consumers, reduced interest rates for marginal borrowers, stable default rates despite expanded access, and improved financial inclusion metrics.

## Stakeholder Incentive Alignment

### Economic Benefits for Lenders

The proposed framework could create multiple value streams for participating lenders. Regulatory capital relief might provide powerful adoption incentive—alternative scoring demonstrating superior performance could potentially qualify for reduced risk weights under internal ratings-based approaches, possibly improving return on equity. Market expansion opportunities through serving previously excluded segments could offset implementation costs. The Buy Now Pay Later sector's rapid growth demonstrates commercial viability of alternative scoring approaches, expanding the addressable market substantially.

Operational efficiencies might reduce per-transaction costs through automated decisioning, straight-through processing, and digital distribution. These efficiency gains could enable profitable service to lower-income segments previously uneconomical to serve.

### Data Provider Value Proposition

Proposed revenue sharing agreements could ensure fair value distribution to data providers contributing to the ecosystem. Telecommunications companies, utilities, and other data custodians might receive transaction fees or basis point shares on originated loans, potentially creating sustainable revenue streams aligned with data quality and availability.

Network effects could increase platform value as participation grows. More lenders might create demand for data, increasing query volumes and revenue. Improved credit access could increase customer lifetime value through reduced churn and enhanced service uptake. Financial inclusion initiatives might enhance brand reputation and corporate social responsibility profiles.

Privacy-preserving technologies could address competitive concerns about data sharing. Differential privacy, secure multi-party computation (allowing joint analysis without sharing raw data), and homomorphic encryption might maintain competitive advantage while enabling collaboration.

### Consumer Benefits and Protections

Granular consent control could maintain trust through transparency. Consumers might select specific data sources, purposes, and duration. Consent dashboards could show active permissions enabling easy revocation. Data portability rights might prevent lock-in. These features would align with global best practices in data governance.

Credit building pathways could incentivize responsible financial behavior. Successful alternative scoring might graduate to bureau reporting after demonstration periods. Positive payment history could transfer across lenders. Credit score improvements might unlock better rates and higher limits.

Protection mechanisms could ensure consumer welfare throughout the process. Mandatory cooling-off periods might prevent over-borrowing. Debt counseling services could provide support. Hardship programs might offer payment relief. Insurance products could protect against income shocks.

## Addressing Implementation Challenges and Risks

### Overcoming Commercial Inertia

Major banks with established credit scoring models may resist adoption due to sunk costs in existing infrastructure and concerns about cannibalizing their current business. To address this, the framework emphasizes regulatory capital relief as a powerful incentive—banks adopting approved alternative scoring models could potentially benefit from reduced risk weights, directly improving return on equity. Additionally, the untapped market potential of 16 million excluded adults represents new revenue opportunities rather than competition for existing customers. Pilot results from the sandbox phase demonstrating maintained portfolio quality while expanding access could provide empirical evidence to overcome skepticism.

### Preventing Data Hoarding

Telecommunications providers and utilities might hesitate to share data due to competitive fears and uncertainty about value capture. The framework addresses this through multiple mechanisms. Privacy-preserving technologies like secure multi-party computation enable data utilization without exposing raw information to competitors. Revenue-sharing models ensure data providers capture fair value from their contribution. Regulatory requirements could mandate participation for systemically important data holders, similar to open banking regulations in other jurisdictions. The success of Kenya's M-Shwari, where Safaricom benefits from both data provision and expanded customer engagement, provides a compelling precedent.

### Building Regulatory Capacity

The NCR may initially lack technical expertise to effectively oversee complex algorithmic systems. The framework proposes a phased capacity-building approach. Initial recruitment of data scientists and machine learning engineers from industry and academia would establish core competency. Partnerships with universities could provide ongoing training and research support. International secondments to regulators with established algorithmic oversight capabilities (such as the UK's Financial Conduct Authority) could accelerate knowledge transfer. Regulatory technology adoption enabling automated monitoring would allow efficient supervision despite resource constraints. The sandbox phase itself serves as a learning laboratory for regulators, building expertise through practical experience before full-scale deployment.

### Managing Algorithmic Bias Risks

Despite best efforts, alternative scoring models might perpetuate or amplify existing societal biases. The framework implements multiple safeguards: mandatory bias audits before deployment and at regular intervals, published fairness metrics ensuring transparency and accountability, diverse training data requirements preventing single-source bias, and explainability standards enabling identification of discriminatory patterns. The three-tier appeals process provides recourse for individuals experiencing potentially discriminatory outcomes. Regular stakeholder forums including civil society organizations would provide external oversight and early warning of emerging issues.

## Regulatory Evolution Requirements

The National Credit Act might require targeted amendments recognizing alternative scoring while maintaining consumer protection. New definitions could distinguish "alternative credit data" from traditional bureau information. Amendments might explicitly permit non-financial data use subject to consent and establish alternative scoring standards including accuracy, fairness, and explainability requirements.

POPIA alignment could necessitate sectoral codes clarifying permissible processing. Legitimate interest grounds might cover financial inclusion objectives. Consent mechanisms could accommodate low-literacy populations through visual and audio formats. Data retention periods might reflect credit lifecycle needs while minimizing privacy risks.

Technical expertise development could ensure effective oversight of algorithmic systems. Recruiting data scientists and ML engineers into regulatory bodies might bring necessary skills. Partnership with academic institutions could provide ongoing training. International secondments might transfer knowledge. Regulatory technology adoption could enable efficient supervision at scale through automated monitoring and pattern analysis.

# Conclusions

This dissertation has developed a conceptual framework for implementing alternative credit scoring in South Africa, addressing the critical challenge of financial exclusion affecting millions of adults. Through systematic literature synthesis and comparative case analysis, the research proposes actionable guidance for expanding credit access while maintaining consumer protection and commercial viability.

The proposed hybrid federated-centralized architecture suggests a path to resolve fundamental tensions between efficiency and privacy, potentially enabling secure data aggregation while respecting POPIA requirements and competitive sensitivities. Edge processing and privacy-preserving technologies could ensure individual privacy while enabling statistical insights. This technical foundation might provide scalable infrastructure supporting the rapidly expanding alternative credit market, as evidenced by the BNPL sector's substantial growth trajectory.

The governance framework proposes clear oversight mechanisms balancing innovation with protection. The suggested Alternative Credit Scoring Authority could provide specialized regulatory capacity while industry self-regulation might enable responsive standard-setting. Three-tier appeals processes could ensure recourse for adverse decisions while behavioral nudges might guide creditworthiness improvement. Mandatory explainability requirements and regular bias audits could maintain algorithmic accountability.

The implementation roadmap suggests a pragmatic pathway from current exclusion to inclusive access. Phase 1 sandbox testing could validate technical and commercial feasibility. Phase 2 scaling might develop operational maturity and market education. Phase 3 ecosystem integration could establish alternative scoring as mainstream infrastructure. This graduated approach, proven successful internationally, could manage risk while building stakeholder confidence.

The framework's economic analysis suggests aligned incentives supporting sustainable adoption. Regulatory capital relief might motivate lender participation. Revenue sharing could incentivize data providers. Credit building pathways might benefit consumers. These mutual benefits could create positive feedback loops driving ecosystem development beyond initial implementation.

Several limitations bound this conceptual contribution. The desk-based methodology precludes primary stakeholder input that might reveal implementation nuances. Rapid technological evolution may alter feasibility assessments. South Africa's specific context limits direct generalizability to other markets. The framework remains untested, requiring empirical validation through pilot implementations. However, the framework's modular structure enables adaptation as conditions evolve and evidence accumulates.

Future research should empirically validate framework propositions through pilot implementations testing technical feasibility and commercial viability. Longitudinal studies tracking inclusion outcomes would assess actual social impact beyond theoretical projections. Comparative analysis across African markets could identify transferable insights and regional adaptation requirements. Technical research on privacy-preserving machine learning might resolve current trade-offs between data utility and privacy protection.

The stakes transcend technical implementation to fundamental questions of economic justice. In a society addressing historical inequalities, financial inclusion represents more than market efficiency—it embodies citizenship, dignity, and opportunity. Alternative credit scoring offers powerful tools for expanding access, but tools alone don't determine outcomes. Success requires commitment to using these capabilities for genuine empowerment rather than sophisticated exclusion. This dissertation provides a conceptual roadmap; the journey requires collective will to build more inclusive financial futures for all South Africans.

---

\newpage

# References